Extended Abstract

### 2021-07-27 paper/paper.tex
Add more interesting typing rules
Rewrite most of the constraint collection section

\usepackage[inline]{enumitem}

@@

\newcommand{\picalc}{$\pi$-calculus}
\newcommand{\Picalc}{$\pi$-Calculus}
\newcommand{\rulename}[1]{{\tiny \textsc{(#1)}}}
\newcommand{\constr}[1]{\textcolor{olive}{\mathtt{#1}}}
\newcommand{\func}[1]{\textcolor{gray}{\mathtt{#1}}}
\newcommand{\type}[1]{\textcolor{blue}{\mathtt{#1}}}

@@

% Typing judgment constructors
\newcommand{\tmvar}{\constr{mvar}}
\newcommand{\tchan}{\constr{chan}}
\newcommand{\tunit}{\constr{unit}}
\newcommand{\tsum}{\constr{sum}}
\newcommand{\tprod}{\constr{prod}}
\newcommand{\tzero}{\constr{0\cdot}}
\newcommand{\tone}{\constr{1\cdot}}
\newcommand{\tomega}{\constr{\omega\cdot}}

@@

% Constraint constructors
\newcommand{\eqconstr}[2]{\constr{[} ~ #1 ~ \constr{\stackrel{c}{=}} ~ #2 ~ \constr{]}}
\newcommand{\sumconstr}[3]{\constr{[} ~ #1 ~ \constr{\stackrel{c}{=}} ~ #2 ~ \constr{+} ~ #3 ~ \constr{]}}

@@

To type check a \picalc{} process with shared and linear types we must assign a type to every communication channel created within the process.
To do so we can either ask the user for type annotations \cite{ZalakainD21}, or we can infer types by looking at how channels are used.
To reduce the burden on the user, we follow the latter approach: we traverse processes bottom-up while keeping a typing context with \emph{metavariables} (\emph{holes}) and collecting typing constraints on metavariables, in line with co-contextual typing \cite{ErdwegBKKM15}.
Keeping a strictly bottom-up information flow has the additional advantage of making typing inference easily parallelisable.
Constraints with a most general solution are solved into substitutions and applied to the typing context, constraints without must be kept around for later: applying them as substitutions risks over-constraining the problem down the line.
Armed with a typing inference algorithm that for a process $P$ infers the most general typing context $\Gamma$ and some typing constraints, type checking that $\Delta \vdash P$ for some $\Delta$ amounts to emitting the extra constraint $\Gamma = \Delta$, assuming the metavariables in $\Gamma$ and $\Delta$ are disjoint.

To the best of our knowledge, this problem has only been treated in Padovani's work on type reconstruction for composite types \cite{Padovani15}.
However, his work does not appear to be mechanised, and metatheoretical properties like soundness and completeness are only informally addressed.
With this work, we aim to:
\begin{itemize}
  \item State and prove clear soundness and completeness theorems for both constraint collection (\autoref{constraint-collection}) and constraint resolution (\autoref{constraint-resolution}).
  \item Mechanise in Agda the totality of this work.
\end{itemize}

We start defining an untyped but well scoped \picalc{} using type-level de Bruijn indices \cite{deBruijn72} and embed a small expression language that handles composite sum and product types.
On top, we define a standard type system with linear and shared types using context-splits \autoref{type-system}.
We then provide an overview of how typing inference works, define constraint collection (\autoref{constraint-collection}) and constraint resolution (\autoref{constraint-resolution}), and state that both phases are sound and complete with regards to the type system defined in \autoref{type-system}.
(A note on notation: variables are black, \textcolor{blue}{types are blue}, \textcolor{olive}{constructors are green}, \textcolor{gray}{functions are gray}, and we are sorry this is not a poem.)

@@

Both types and usage annotations contain metavariables.
To make their handling uniform we use a common set of metavariables for both.
A context of kinds \(\gamma\) keeps track of whether a metavariable is of type kind $\ktype$ or usage annotation kind $\kusage$.
Variable references $m$ are of type $\tvar{\gamma}{k}$: the set of variables of kind \(k\) in a kinding context \(\gamma\).
Usage annotations $i$ and $o$ are of type $\tkind{\gamma}{\kusage}$, and types $s$ and $t$ of type $\tkind{\gamma}{\ktype}$.
We henceforth abbreviate $\tkind{\gamma}{\ktype}$ as $\ttype{\gamma}$ and $\tkind{\gamma}{\kusage}$ as $\tusage{\gamma}$.
\[
\begin{aligned}[c]
  i ~ o :=& ~ \tmvar_{\kusage} ~ m ~|~ \tzero ~|~ \tone ~|~ \tomega \\
  s ~ t :=& ~ \tmvar_{\ktype} ~ m ~|~ \tunit ~|~ \tchan ~ i ~ o ~ t ~|~ \tprod ~ s ~ t ~|~ \tsum ~ s ~ t
\end{aligned}
\]

@@

  \inferrule {\tSplit{i_x}{i_y}{i_z} \\ \tSplit{o_x}{o_y}{o_z}} {\tSplit{\tchan~i_x~o_x~t}{\tchan~i_y~o_y~t}{\tchan~i_z~o_z~t}}
\end{mathpar}
Following \cite{Padovani15}, a context $\Gamma$ is unrestricted (non-linear) $\tun{\Gamma}$ if it can be split into itself: $\tSplit{\Gamma}{\Gamma}{\Gamma}$ (respectively $\tun{x}$ and $\tun{t}$ for usage annotations $x$ and types $t$).

\paragraph{Typing Judgments}
\label{typing-judgments}
We can now define our typing judgment for variables ($\tVar{\Gamma}{i}{t}$, where variable $i$ under context $\Gamma$ is of type $t$), expressions ($\tExpr{\Gamma}{e}{t}$, where expression $e$ under context $\Gamma$ is well typed with type $t$) and processes ($\tProc{\Gamma}{p}$, where process $p$ is well typed under context $\Gamma$). Some cases are omitted for brevity.

\begin{mathpar}
%  \inferrule { \Gamma : \tCtx{n}{\gamma} \\ i : \Fin{n} \\ t : \ttype{\gamma} }
%             { \tVar{\Gamma}{i}{t} }
  \inferrule { \Gamma : \tCtx{n}{\gamma} \\ \tun{\Gamma} \\ t : \ttype{\gamma} }
             { \tVar{\Gamma,t}{\fzero}{t} }
             % \rulename{zero}

  \inferrule { \tVar{\Gamma}{i}{t} \\ s : \ttype{\gamma} \\ \tun{s} }
             { \tVar{\Gamma,s}{\fsuc~i}{t} }
             % \rulename{suc}
%  \inferrule { \Gamma : \tCtx{n}{\gamma} \\ e : \sExpr{n} \\ t : \ttype{\gamma}}
%             { \tExpr{\Gamma}{e}{t} }

  \inferrule { \tVar{\Gamma}{i}{t} } { \tExpr{\Gamma}{\svar~i}{t} }
             % \rulename{var}
  %\inferrule { \tun{\Gamma} } { \tExpr{\Gamma}{\sunit}{\tunit} }
             % \rulename{unit}

  \inferrule { \tExpr{\Gamma}{e}{\tprod~t~s} \\ \tun{s}} {\tExpr{\Gamma}{\sfst~e}{t}}
             % \rulename{fst}

  \inferrule { \tExpr{\Gamma}{e}{t} } { \tExpr{\Gamma}{\sinl~e}{\tsum~t~s}}
             % \rulename{inl}
  %\inferrule { \tSplit{\Gamma}{\Delta}{\Theta} \\ \tExpr{\Delta}{e}{s} \\ \tExpr{\Theta}{f}{t} }
  %           { \tExpr{\Gamma}{\spair~e~f}{\tprod~s~t}}
             %\rulename{pair}
%  \inferrule { \Gamma : \tCtx{n}{\gamma} \\ p : \sProc{n}}
%             { \tProc{\Gamma}{p} }

  \inferrule { \tun{\Gamma} } { \tProc{\Gamma}{\send} }
             %\rulename{end}

  \inferrule { \tProc{\Gamma}{p} \\ \tun{\Gamma} }
             { \tProc{\Gamma}{\srec~p} }
             %\rulename{rec}

  \inferrule { t : \ttype{\gamma} \\ \tProc{\Gamma , t}{p} } { \tProc{\Gamma}{\snew~p} }
             %\rulename{new}
  %\inferrule { \tSplit{\Gamma}{\Delta}{\Theta} \\ \tExpr{\Delta}{e}{\tchan~\tone~\tzero~t} \\ \tProc{\Theta,t}{p} }
  %           { \tProc{\Gamma}{\srecv~e~p} }
  %           %\rulename{recv}

  \inferrule { \tSplit{\Gamma}{\Delta}{\Theta} \\ \tExpr{\Delta}{e}{\tsum~s~t} \\ \tProc{\Theta,s}{p} \\ \tProc{\Theta,t}{q} }
             { \tProc{\Gamma}{\scase~e~p~q} }
             %\rulename{case}
\end{mathpar}

@@

Co-contextual typing inference traverses processes bottom-up keeping a typing context and collecting constraints.
This makes inference subproblems independent and thus easily parallelisable.
In the shared \picalc{}, inference constraints always have a most general unifier, in the shared \emph{and} linear \picalc{} however they do not.
Consider the open process $\ssend~a~\sunit~(\ssend~x~a~\send)$ where $x$ and $a$ are free: we partly use $a$ to send, then send $a$ away over $x$ and terminate.
When the process terminates, all usage annotations must be unrestricted, but we cannot decide whether to instantiate with $\tzero$ or $\tomega$.
Similarly, we ignore \emph{how much} of $a$ we are sending away over $x$: that depends on what the receiving process expects.
Nonetheless, we do have to keep track of the fact that $a$ has been used once to send.

In \autoref{constraint-collection} we introduce typing constraints, and define kind-preserving substitution, constraint satisfaction, and a typing inference algorithm.
We show that satisfying the constraints generated by typing inference is enough to make a process typable (soundness), and postulate that for any typable process typing inference will find the most general set of constraints that makes the process typable (completeness).

In \autoref{constraint-resolution} we solve equality constraints by unification and context split constraints by decomposition.
Constraints with a most general solution can be solved into substitutions.
(Constraints without a most general solution can be solved by instantiation once the process is closed.)
Solving a set of constraints thus results in a set of substitutions and a set of simplified constraints that had those substitutions applied.
We postulate that satisfying the simplified constraints amounts to satisfying the original constraints after substitution (soundness), and that for every substitution that solves the original constraints we can find a most general substitution that will solve the simplified constraints (completeness).
Additionally, we postulate that the simplified set of constraints only contains constraints without a most general solution (progress).

\subsection{Constraint Collection}
\label{constraint-collection}

@@

Constraints of type $\tConstr{\gamma}$ are defined on arguments of type $\tkind{\gamma}{k}$ for some $k$ --- that is, on both usage annotations and types.
They take two forms: the binary $\eqconstr{S}{T}$, where $S$ and $T$ must be unified, and the ternary $\sumconstr{S}{T}{R}$, where $T$ and $R$ must add up to $S$.
Constraints of the former form can \emph{always} be eagerly solved, while constraints of the latter can \emph{sometimes} not.
We use $\tConstrs{\gamma}$ to refer to lists of constraints of type $\tConstr{\gamma}$.

@@

We use a $\interpr{\_}$ function to interpret constraints $\eqconstr{S}{T}$ and $\sumconstr{S}{T}{R}$ into their reified counterparts $\tEq{S}{T}$ and $\tSplit{S}{T}{R}$, respectively.

@@

Typing inference takes a process with $n$ free variables and returns a metavariable context $\gamma$, a typing context with $n$ free variables containing metavariables in $\gamma$, and a list of constraints on metavariables $\gamma$.
We define a similar function for typing inference on expressions, this time also returning a type in $\ttype{\gamma}$:
\begin{flalign*}
& \func{inferProc} : \sProc{n} \to ~ \exists \gamma. ~ \tCtx{n}{\gamma} \times \tConstrs{\gamma} && \\
& \func{inferExpr} : \sExpr{n} \to ~ \exists \gamma. ~ \tCtx{n}{\gamma} \times \tConstrs{\gamma} \times \ttype{\gamma} &&
\end{flalign*}
These functions are total: if a process or expression is untypable its constraints will be unsolvable.

Let us provide a couple of examples illustrating how this function is defined:
For the expression $\sfst~e$ we
\begin{enumerate*}[label=\textcolor{gray}{\arabic*)}]
  \item recursively infer $e$ resulting in $\gamma,\Gamma,Cs,\tau$;
  \item add two fresh type metavariables $?t$ and $?s$ to $\gamma$;
  \item add constraints $\eqconstr{\tau}{\tprod~?t~?s}$ and $\sumconstr{?s}{?s}{?s}$ to $Cs$; and
  \item return $?t$ as the type of $\sfst~e$.
\end{enumerate*}

For the expression $\scase~e~p~q$ we
  \begin{enumerate*}[label=\textcolor{gray}{\arabic*)}]
  \item infer $p$ resulting in $(\theta_p, \Theta_p, s , Cs_p)$;
  \item infer $q$ resulting in $(\theta_q, \Theta_q, t , Cs_q)$;
  \item infer $e$ resulting in $(\delta_e, \Delta_e, Cs_e, \tau)$;
  \item create a new metavariable context $\gamma$ and a new typing context $\Gamma$ where every type is a fresh metavariable in $\gamma$;
  \item take the union of the metavariables in $\gamma$, $\theta_p$, $\theta_q$ and $\delta_e$;
  \item take the union of the constraints in $Cs_p$, $Cs_q$ and $Cs_e$, and create new constraints $\eqconstr{\Theta_p}{\Theta_q}$, $\sumconstr{\Gamma}{\Delta_e}{\Theta_p}$, and $\eqconstr{\tau}{\tsum~s~t}$; and
  \item return $\Gamma$ as the inferred context.
\end{enumerate*}

\begin{mytheorem}[Inference Soundness]\label{inference-soundness}
Given $\func{infer}~p$ returns $\gamma , \Gamma , cs$, every substitution \(\sigma\) that makes the constraints $cs$ hold makes $\tProc{(\subst{\sigma}{\Gamma})}{p}$ hold.
\end{mytheorem}

\begin{mytheorem}[Inference Completeness]\label{inference-completeness}
Given $\func{infer}~p$ returns $\gamma , \Gamma , cs$, for every context \(\Delta\) that makes $p$ typable, there exists a substitution \(\sigma\) that solves the constraints $cs$ such that $\Delta$ is a specialisation of $(\subst{\sigma}{\Gamma})$ --- a specialisation $\Delta ~ \func{\subseteq} ~ (\subst{\sigma}{\Gamma})$ is defined as $\exists \sigma_f . ~ \tEq{\Delta}{(\subst{\sigma_f}{(\subst{\sigma}{\Gamma})})}$.
\end{mytheorem}

>>>>

\usepackage[inline]{enumitem}
\usepackage{todonotes}

@@

\newcommand{\picalc}{$\pi$-calculus}
\newcommand{\Picalc}{$\pi$-Calculus}
\newcommand{\rulename}[1]{{\tiny \textsc{(#1)}}}
\newcommand{\var}[1]{\textcolor{black}{\mathnormal{#1}}}
\newcommand{\constr}[1]{\textcolor{olive}{\mathtt{#1}}}
\newcommand{\func}[1]{\textcolor{gray}{\mathtt{#1}}}
\newcommand{\type}[1]{\textcolor{blue}{\mathtt{#1}}}

@@

% Typing judgment constructors
\newcommand{\tmvar}{\constr{mvar}}
\newcommand{\tchan}[3]{\constr{chan}_{\constr{[}#1\constr{,}#2\constr{]}} ~ #3}
\newcommand{\tunit}{\constr{unit}}
\newcommand{\tsum}{\constr{sum}}
\newcommand{\tprod}{\constr{prod}}
\newcommand{\tzero}{\constr{0\cdot}}
\newcommand{\tone}{\constr{1\cdot}}
\newcommand{\tomega}{\constr{\omega\cdot}}

@@

% Constraint constructors
\newcommand{\eqconstr}[2]{\constr{[} ~ #1 ~ \constr{\stackrel{c}{=}} ~ #2 ~ \constr{]}}
\newcommand{\sumconstr}[3]{\constr{[} ~ #1 ~ \constr{\stackrel{c}{=}} ~ #2 ~ \constr{+} ~ #3 ~ \constr{]}}
\newcommand{\varconstr}[3]{\constr{[} ~ #1 ~ \constr{\stackrel{c}{\ni}_{\var{#2}}} ~ #3 ~ \constr{]}}

@@

To type check a \picalc{} process with shared and linear types we must assign a type to every communication channel created within the process.
To do so we can either ask the user for type annotations \cite{ZalakainD21}, or we can infer types by looking at how channels are used.
To reduce the burden on the user, we follow the latter approach: we traverse processes bottom-up while keeping a typing context with \emph{metavariables} (\emph{holes}) and collecting typing constraints on metavariables, in line with co-contextual typing \cite{ErdwegBKKM15}.
Keeping a strictly bottom-up information flow has the additional advantage of making typing inference parallelisable.
Constraints with a most general solution are solved into substitutions and applied to the typing context, constraints without must be kept around for later: applying them as substitutions risks over-constraining the problem down the line.
Armed with a typing inference algorithm that for a process $P$ infers the most general typing context $\Gamma$ and some typing constraints, type checking that $\Delta \vdash P$ for some $\Delta$ amounts to emitting the extra constraint $\Gamma = \Delta$, assuming the metavariables in $\Gamma$ and $\Delta$ are disjoint.

To the best of our knowledge, this problem has only been treated in Padovani's work on type reconstruction for composite types \cite{Padovani15}.
However, his work does not appear to be mechanised, and metatheoretical properties like soundness and completeness are only informally addressed.
With this work, we aim to:
\begin{itemize}
  \item State and prove clear soundness and completeness theorems for both constraint collection and constraint resolution (\autoref{inference}).
  \item Mechanise in Agda the totality of this work.
\end{itemize}

We start defining an untyped but well scoped \picalc{} using type-level de Bruijn indices \cite{deBruijn72} and embed a small expression language that handles composite sum and product types.
On top, we define a standard type system with linear and shared types using context-splits (\autoref{type-system}).
We then provide an overview of how typing inference works, define constraint collection and constraint resolution (\autoref{inference}), and state that both phases are sound and complete with regards to the type system defined in \autoref{type-system}.
(Notation: variables are black, \textcolor{blue}{types are blue}, \textcolor{olive}{constructors are green}, and \textcolor{gray}{functions are gray}.)

@@

Both types and usage annotations contain metavariables.
To make their handling uniform we use a common set of metavariables for both.
A context of kinds \(\gamma\) keeps track of whether a metavariable is of type kind $\ktype$ or usage annotation kind $\kusage$.
Variable references $m$ are of type $\tvar{\gamma}{k}$: the set of variables of kind \(k\) in a kinding context \(\gamma\).
Usage annotations $i$ and $o$ are of type $\tkind{\gamma}{\kusage}$, and types $s$ and $t$ of type $\tkind{\gamma}{\ktype}$.
We henceforth abbreviate $\tkind{\gamma}{\ktype}$ as $\ttype{\gamma}$ and $\tkind{\gamma}{\kusage}$ as $\tusage{\gamma}$.
\[
\begin{aligned}[c]
  i ~ o :=& ~ \tmvar_{\kusage} ~ m ~|~ \tzero ~|~ \tone ~|~ \tomega \\
  s ~ t :=& ~ \tmvar_{\ktype} ~ m ~|~ \tunit ~|~ \tchan{i}{o}{t} ~|~ \tprod ~ s ~ t ~|~ \tsum ~ s ~ t
\end{aligned}
\]

@@

  \inferrule {\tSplit{i_x}{i_y}{i_z} \\ \tSplit{o_x}{o_y}{o_z}} {\tSplit{\tchan{i_x}{o_x}{t}}{\tchan{i_y}{o_y}{t}}{\tchan{i_z}{o_z}{t}}}
\end{mathpar}
Following \cite{Padovani15}, a context $\Gamma$ is unrestricted (non-linear) $\tun{\Gamma}$ if it can be split into itself: $\tSplit{\Gamma}{\Gamma}{\Gamma}$ (respectively $\tun{x}$ and $\tun{t}$ for usage annotations $x$ and types $t$).

\paragraph{Typing Judgments}
\label{typing-judgments}
We can now define our typing judgment for variables ($\tVar{\Gamma}{i}{t}$, where variable $i$ under context $\Gamma$ is of type $t$), expressions ($\tExpr{\Gamma}{e}{t}$, where expression $e$ under context $\Gamma$ is well typed with type $t$) and processes ($\tProc{\Gamma}{p}$, where process $p$ is well typed under context $\Gamma$).
Note that, although fully linear tensor products ($\tprod ~ s ~ t$) cannot directly be eliminated through their projections ($\sfst$ rule), context splits enable their usage annotations to be distributed before projection.
(Some cases are omitted for brevity.)

\begin{mathpar}
  \inferrule { \Gamma : \tCtx{n}{\gamma} \\ \tun{\Gamma} \\ t : \ttype{\gamma} }
             { \tVar{\Gamma,t}{\fzero}{t} }

  \inferrule { \tVar{\Gamma}{i}{t} \\ s : \ttype{\gamma} \\ \tun{s} }
             { \tVar{\Gamma,s}{\fsuc~i}{t} }

  \inferrule { \tExpr{\Gamma}{e}{\tprod~t~s} \\ \tun{s}} {\tExpr{\Gamma}{\sfst~e}{t}}

  \inferrule { \tSplit{\Gamma}{\Delta}{\Theta} \\ \tExpr{\Delta}{e}{s} \\ \tExpr{\Theta}{f}{t} }
             { \tExpr{\Gamma}{\spair~e~f}{\tprod~s~t}}

  \inferrule { \tun{\Gamma} } { \tProc{\Gamma}{\send} }

  \inferrule { t : \ttype{\gamma} \\ \tProc{\Gamma , t}{p} } { \tProc{\Gamma}{\snew~p} }

  \inferrule { \tSplit{\Gamma}{\Delta}{\Theta} \\ \tExpr{\Delta}{e}{\tchan{\tone}{\tzero}{t}} \\ \tProc{\Theta,t}{p} }
             { \tProc{\Gamma}{\srecv~e~p} }

  \mprset{flushleft}
  \inferrule {
    \tSplit{\Gamma}{\Delta}{\Theta} \\\\
    \tSplit{\Theta}{\Psi}{\Xi} \\
    \tExpr{\Delta}{f}{\tchan{\tzero}{\tone}{t}} \\
    \tExpr{\Psi}{e}{t} \\
    \tProc{\Xi}{p}
  }
  { \tProc{\Gamma}{\ssend~f~e~p} }
\end{mathpar}

@@

@@

Constraints of type $\tConstr{\gamma}$ are defined on arguments of type $\tkind{\gamma}{k}$ for some $k$ --- that is, on both usage annotations and types.
They take two forms: the binary $\eqconstr{S}{T}$, where $S$ and $T$ must be unified, and the ternary $\sumconstr{S}{T}{R}$, where $S$ is split into $T$ and $R$.
Constraints are pointwise lifted to typing contexts, and we abbreviate with $\varconstr{\Gamma}{i}{t}$ a constraint that states that $i$ must be of type $t$ in $\Gamma$, and all other variables in $\Gamma$ must be unrestricted.
We use $\tConstrs{\gamma}$ to refer to lists of constraints of type $\tConstr{\gamma}$.

@@

We use a $\interpr{\_}$ function to interpret constraints $\eqconstr{S}{T}$ and $\sumconstr{S}{T}{R}$ into claims of their satisfiability $\tEq{S}{T}$ and $\tSplit{S}{T}{R}$, respectively.

@@

Typing inference takes a process with $n$ free variables and returns a metavariable context $\gamma$, a typing context with $n$ free variables containing metavariables in $\gamma$, and a list of constraints on metavariables $\gamma$.
We define a similar function for typing inference on expressions, this time also returning a type in $\ttype{\gamma}$.
These functions are total: if a process or expression is untypable its constraints are unsolvable.
\begin{flalign*}
& \func{inferProc} : \sProc{n} \to ~ \exists \gamma. ~ \tCtx{n}{\gamma} \times \tConstrs{\gamma} && \\
& \func{inferExpr} : \sExpr{n} \to ~ \exists \gamma. ~ \tCtx{n}{\gamma} \times \tConstrs{\gamma} \times \ttype{\gamma} &&
\end{flalign*}

The reason we return unsolved constraints is that unlike in the shared \picalc{} where constraints have always a most general unifier, in the shared \emph{and} linear \picalc{} metavariables can be under-constrained. 
Consider the open process $\ssend~a~\sunit~(\ssend~x~a~\send)$ where $x$ and $a$ are free: we partly use $a$ to send, then send whatever is left of $a$ away over $x$ and terminate.
Let us step through a working example of how inference runs:
\begin{enumerate}
  \item on $\send$, inference creates a fresh typing context $\Gamma_0$ with metavariables for $x$ and $a$, and constraints demanding that these should be unrestricted.
  \item on $\ssend ~ x ~ a$, inference creates fresh typing contexts $\Gamma_4$, $\Gamma_3$, $\Gamma_2$ and $\Gamma_1$, a fresh metavariable $?t$, and constraints $\sumconstr{\Gamma_4}{\Gamma_3}{\Gamma_2}$, $\varconstr{\Gamma_3}{x}{\tchan{\tzero}{\tone}{?t}}$, $\sumconstr{\Gamma_2}{\Gamma_1}{\Gamma_0}$, and $\varconstr{\Gamma_1}{a}{?t}$.
  \item on $\ssend ~ a ~ \sunit$, inference creates fresh typing contexts $\Gamma_6$, and $\Gamma_5$, and constraints $\sumconstr{\Gamma_6}{\Gamma_5}{\Gamma_4}$, and $\varconstr{\Gamma_5}{a}{\tchan{\tzero}{\tone}{\tunit}}$.
\end{enumerate}

Two reasons prevent us from finding most general solutions:
On one hand, \emph{unrestrictedness} constraints are satisfied by either usage annotations $\tzero$ and $\tomega$.
This can be overcome by a subtyping relation on usage annotations.
Additionally, \emph{usage polymorphism} on $?t$ makes inference under-constrained.
The overall constraint on the metavariable for $a$ is that it can be split into $\tchan{\tzero}{\tone}{\tunit}$, into $?t$, and into some unrestricted leftovers.
While $?t$ can certainly be eagerly substituted by a channel type $\tchan{?i}{?o}{\tunit}$ for some $?i$ and $?o$, we cannot determine what $?i$ and $?o$ must be.
The problem here is that we must keep track of the partial usage of $a$ while at the same time allowing the type of $a$ to be polymorphic in the type of $x$.
As a result, we must keep these under-constrained constraints around (potentially until the process is closed and they can be solved by instantiation) and state meta-theoretical properties abstracting over constraint satisfaction.

\begin{mytheorem}[Inference Soundness]\label{inference-soundness}
Given $\func{infer}~p$ returns $(\gamma , \Gamma , cs)$, every substitution \(\sigma\) satisfying $\interpr{\subst{\sigma}{cs}}$ makes $\tProc{(\subst{\sigma}{\Gamma})}{p}$ hold.
\end{mytheorem}

\begin{mytheorem}[Inference Completeness]\label{inference-completeness}
Given $\func{infer}~p$ returns $(\gamma , \Gamma , cs)$, for every context \(\Delta\) such that $\tProc{\Delta}{p}$, there exists a substitution \(\sigma\) satisfying $\interpr{\subst{\sigma}{cs}}$ such that $\Delta$ is a specialisation of $(\subst{\sigma}{\Gamma})$ --- a specialisation $\Delta ~ \func{\subseteq} ~ (\subst{\sigma}{\Gamma})$ is defined as $\exists \sigma_f . ~ \tEq{\Delta}{(\subst{\sigma_f}{(\subst{\sigma}{\Gamma})})}$.
\end{mytheorem}
